---
title: "Midterm STT 5811"
author: "Your Name"
date: '`r Sys.Date()`'
output: html_document
---

```{r, label = "SETUP", echo = FALSE, results= 'hide', message = FALSE, warning = FALSE}
library(knitr)
library(stringi)
knitr::opts_chunk$set(comment = NA, fig.show = 'as.is', fig.height = 4, fig.width = 4, prompt = TRUE, highlight = TRUE, tidy = FALSE, warning = FALSE, message = FALSE, tidy.opts=list(blank = TRUE, width.cutoff= 75, cache = TRUE))
GA <- read.csv("GA.csv", colClasses = "character")
```


**Midterm Directions:**  Answer all questions to the midterm with complete sentences.  Make sure to show all code and use inline code when reporting numerical answers to questions.  The Gettysburg address is stored in the `Data` folder of the class repository as `GA.csv`.  Please italicize all answers. Create a directory under your private GitHub repository named `MidTerm`.  Move this document (`MidTerm5811.Rmd`) to your `MidTerm` directory as well as the `GA.csv` file.  Push your completed work to your private repository no later than 17:00 October 9, 2015.   


(a)  Read the Gettysburg address into an object named `GA` using the `read.csv` function (use `colClasses = "character"`).  Convert all words used in the Gettysburg address to lower case using the function `tolower`.  Show the first six words in the object `GA`. [2 pts]  

```{r}
# Your code
```

(b) Set the seed value to 649 (`set.seed(649)`).  Use the function `sample` to take a simple random sample of size 10 from the Gettysburg address and store the results in the object `SAMPLE10`.  Show the results of `SAMPLE10`. [2 pts]

```{r}
set.seed(649)
# Your code
```

(c) Consider the following variables and identify each as quantitative or categorical: [2 pts]
* length of word (number of letters) - Type: *Your answer*
* whether or not word contains the letter `e` - Type: *Your answer*

(d) Use the package `stringi` and write `R` code to determine the string length of each word in `SAMPLE10`. Create a logical vector indicating whether each word in `SAMPLE10` contains one or more `e`s. Use `R` code to count the number of words containing one or more `e`s in `SAMPLE10`. [4 pts]

```{r}
# Your code
```

(e) What is the average word length for `SAMPLE10`?  [2 pts]

```{r}
# Your code
```

* *Your answer*

(f) Is the average length you calculated in d. a parameter or a statistic? Explain. Do we know its value? What symbol can we use to denote its value? [3 pts]

* *Your answer*

* *Your answer*

* *Your answer*

(g) The average length of all 268 words in this population is `r mean(stringi::stri_length(GA$words))` letters. Is this number a parameter or a statistic?  Do we know its value? What symbol can we use to denote its value? [3 pts]

* *Your answer*

* *Your answer*

* *Your answer*

(h) Simulate $m = 150$ students each taking a simple random sample (of size $n = 10$) of the words in the Gettysburg address.  Use a seed of 55.  Store the length of each word in an `R` object (Hint: use a matrix).  Construct a dotplot displaying the average word length of the SRS for each of the the $m = 150$ students. [12 pts]

*This simulation pretends there are $m=150$ students.*

```{r}
set.seed(55)
# Your code
```


(i)  Characterize the distribution of these sample means. [3 pts]


```{r}
# Your code
```

* *Your answer*

(j) Write `R` code that determines how many and what proportion of the $m = 150$ simulated students had a sample mean word length that exceeded the population mean word length? [2 pts]

```{r}
# Your code
```


(k)  Write `R` code that simulates $m = 10000$ students each taking a simple random sample of size $n = 81$ words from the Gettysburg address using a seed of 34.  Create a data frame `DF` with a variable `AvgLen` that contains the average word length for each of the $m = 10000$ simple random samples.  Show the first eight values of `AvgLen`.  [10 pts] 

```{r}
set.seed(34)
# Your code
```


(l) Compute the population mean ($\mu_{\bar{x}}$) and population standard deviation for average word length ($\sigma_{\bar{x}}$) of the Gettysburg address storing the results in the objects `EW` and `SWA`, respectively.  Make sure to use a finite population correction factor, $(N - n)/(N - 1)$, when computing $\sigma_{\bar{x}}$.  Store the value for $\sigma_{\bar{x}}$ without a finite population correction factor in `SW`.  Compute the mean and standard deviation of the values in `AvgLen`. [12 pts]  


```{r}
# Your code
```

(m)  Based on your results, explain whether or not your simulation of $m = 10000$ students each taking a simple random sample of size $n = 81$ can be used to obtain an unbiased estimate of $\mu$. [3 pts]  

* *Your answer*

(n) Create a density plot of the values in `AvgLen`.  Superimpose a normal distribution with a darkgreen line over the density plot using the parameters stored in `EW` and `SWA` for the mean and standard deviation respectively. Superimpose a second normal distribution with a lightblue line over the density plot using the parameters stored in `EW` and `SW` for the mean and standard deviation respectively. Does the darkgreen or lightblue line provide a better model for sampling without replacement?  [10 pts] 

```{r}
# Your code
```

* *Your answer*

(o) Suppose you wanted to achieve sample means that tended to fall even closer to the population mean, would you increase the sample size or the number of samples? [2 pts]

* *Your answer*


(p) Use `set.seed = 145` to select a simple random sample of size ($n = 5$) from the words in the Gettysburg address and store the results in the object `SAMPLE`.  Write `R` code to determine the proportion of the words in `SAMPLE` that contain the letter `e`. [3 pts]

```{r}
set.seed(145)
# Your code
```

* *Your answer*

(q) Construct 95% confidence intervals for $\pi$ the true proportion of words in the Gettysburg address that have at least one `e` using the Wald (asymptotic) and Agresti-Coull methods using the values obtained in (p).  What is the actual value for $\pi$?  Do either or both of the constructed confidence intervals contain $\pi$?  Interpret the constructed confidence intervals. [10 pts]   

```{r}
# Your code
```

*  *Your answer*

(r) Use `set.seed = 3` to select a simple random sample of size ($n = 5$) from the words in the Gettysburg address and store the results in the object `SAMPLE`.  Write `R` code to determine the proportion of the words in `SAMPLE` that contain the letter `e`.   Construct 95% confidence intervals for $\pi$ the true proportion of words in the Gettysburg address that have at least one `e` using the Wald (asymptotic) and Agresti-Coull methods.  Does the interpretation of the constructed confidence intervals change? [10 pts] 

```{r}
set.seed(3)
# Your code
```

*  *Your answer*
*  *Your interpretation*


(s)  Suppose we are interested in testing whether the proportion of `the`s in a large work exceeds 10%.  If the true proportion of `the`s is 20% and we take a random sample of 100 words, what is the power of the test if the critical value used to reject the null hypothesis is 16 or more `the`s? [5 pts]

```{r}
# Your code
```

* *Your answer*

